Using tours to visually investigate properties of new projection pursuit indexes with application to problems in physics

### 2019-01-17 scagnostics.Rmd
Spell checks.

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behavior of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr] provides optimisation routines, and visualization.

@@

The paper investigates the behavior of these indexes. Section \ref{sec:construct} discusses index construction, and how they fit into the guided tour. Section \ref{sec:investigate} investigates the behavior of the indexes, explored primarily using tour methods (@As85, @BCAH05, @CBCH94). The new guided tour with these indexes are applied to two examples from physics models (Section \ref{sec:phys}).

@@

- Standardizing each variable, to have mean 0 and variance 1, so that individual variable scales do not affect the result. Different variable scales are examinable without resorting to projection pursuit, so can be handled prior to searching through high dimensions.
- Sphering the high-dimensional data is often done to remove linear dependence. This is typically done using principal component analysis, and using the principal components as the variables passed to PP. If linear dependence is the only structure PP is not needed, and thus this is removed before PP so that the PPIs are not distracted by simple structure.
- Transform single variables to reduce skewness. It is marginal structure, visible in a single variable, which doesn't need a multivariate technique to reveal. Skewed distributions will inadvertently affect the PPIs, distracting the search for dependence. 
- Remove outliers, which may be an iterative process, to discover, identify and delete. Extreme values will likely affect PPI performance. Outliers can be examined on a case by case basis later.
- Possibly remove noise dimensions, which is also likely to be an iterative process. Directions where the distribution is purely noise make optimization of a PPI more difficult. If a variable is suspected to have little structure and relationship with other variables, conducting PP on the subset of variables without them may improve the efficiency of the search.
- Centering and scaling of the projected data, can be helpful visually. If the data has a small amount of non-normal distribution in some directions, the projected data can appear to wander around the plot window during a tour. It doesn't matter what the center of the projected data is, so centering removes a wandering scatterplot. Less commonly, it may be useful to scale the projected data to standard values, which would be done to remove any linear dependence remaining in the data.

@@

\begin{table}[htp]
\caption{Summary of notation}
\centering
\begin{tabular}{|l|p{8.2cm}|} \hline
$p$ & be the number of variables in the data. For physics models the obervable space is higher dimension, say $q$, and the data examined is the fitted model space, typically less than 10. \\
$d (=2) $ & projection dimension. For studying physics models, typically $d=2$ and this is the focus for the index definition. \\
$n$ & Number of observations, which for the physics models is the number of fitted models being examined and compared. \\
&\\
$\bm{X}_{n\times p} = (\bm{X}_1, ..., \bm{X}_p)$ & $n\times p$-dimensional data matrix, where variables $\bm{X}_j$ may be scaled or standardised, and $\bm{X}$ may be sphered\\
$\bm{Y}_{n\times 2} = (\bm{Y}_1, ..., \bm{Y}_2)$ & projected data matrix. where $Y_j = (\alpha_1 X_1, ..., \alpha_p X_p$ \\
$\bm{F}_{p\times 2} = (\bm{\alpha}_1, ..., \bm{\alpha}_2)$ & orthonomal projection matrix\\ 
&\\
$H$ & Convex hull \\
$A$ & Alpha hull \\\hline
\end{tabular}
\label{notation}
\end{table}

@@

* **scagnostics**: The first step to computing the scagnostics is that the bivariate data is binned, and scaled between [0,1] for calculations. The convex and alpha hulls, and the minimal spanning tree (MST), are computed. 
    + *convex*: The ratio of the area of alpha to convenx hull, $I_{\bm{F}, convex}= \frac{area(A(Y))}{area(H(Y))}$. In this is the only measure where interesting projections will take low values, with a maximum of 1 if both areas are the same. Thus $1-c_{convex}$ is used.
    + *skinny*: The ratio of the perimeter to the area of the alpha hull, $I_{\bm{F}, skinny} = 1 -  \frac{\sqrt{4\pi area(A)}}{perimeter(A)}$, where the normalization is chosen such that $I_{\bm{F}, skinny} = 0$ for a full circle. Values close to 1 indicate a skinny polygon.
    + *stringy*: Based on the MST, $I_{\bm{F}, stringy} = \frac{diameter(MST)}{length(MST)}$ where the diameter is the longest connected path, and the length is the total length (sum of all edges). If the MST contains no branches $I_{\bm{F}, skinny} = 1$.

@@

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuations or a consequence of A noisy index function. \cite{posse95b} discusses the optimization, in particular that for most index functions and optimal results are too local, largely dependent on starting point. @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimization for the identification of the maximum. For exploring high-dimensional data, it can be interesting to observe local maxima as well as a global maximum, and thus a hybrid algorithm that still aloows lingering but not being trapped by local maxima is ideal. In addition, being able to visually monitor the optimisation and see the optimal projection in the context of neighbouring projections is useful. This is provided by combining projection pursuit with the grand tour [@CBCH94]. The properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshooting and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible algorithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. The tourr package also provides the \texttt{search\_geodesic} function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimises the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers, e.g. @diaconis84, @huber85, @JS87, @posse95b, @hall89. The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modeling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimization feasible, small angles between projections correspond to small perturbations of the sample, and thus should be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the criteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.

@@

Three families of data simulations are used for examining the behavior of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the PPIs on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $n$ is used for the number of points, $p$ is the number of dimensions, and $d=2$ is the projection dimension.

@@

Rescaling: first have cutoff at the respective value for the normal distribution, i.e. any value below 0.2 (0.8) is set to zero for cmass (holes) index, rescale remaining range to be between zero and one.

@@

XXX from caption, remove?:  For too rough binning (bins=10) we observe random fluctuations from the binning, with structure detedction being obscured. With bins=20 we see somewhat smoother behaviour for the stringy index, especially when moving into the maximum, while the skinny index shows a dip in the final view, and larger bins=50 is preferred. For the convex index we also find bins=50 to be the better choice.

@@

## Binning sensitivity of MIC index
To examine the sensitivity of binning in the MIC PPI, the classic RANDU data [@Marsaglia68], available in R, is used. Binning is defined by $\delta$, where $B(n) = n^{\delta}$. Figure \ref{fig:randuEx} shows the best projection, index value and computing time obtained when optimizing the MIC index with values $\delta = 0.6, 0.7, 0.8$. With small $\delta$, less bins, the structure isn't visible, and with larger $\delta$ the structure is confounded with noise. It does appear that this parameter affects the performace of the MIC index.

>>>>

To find appropriate projections pursuit indexes, literature on variable selection was examined. With high-dimensional data, even plotting all pairs of variables can lead to many plots, which is what "scagnostics" (@scag, @WW08) were developed to address by providing metrics from which to select the most interesting variable pairs. There are eight scagnostics, of which three ("convex", "skinny" and "stringy") are used here. The question is whether these can be adapted into projection pursuit indexes, to search for unusual features in two-dimensional projections of high-dimensional data. Recent PhD research by @Grimm2016 explored the behaviour of scagnostics for selecting variables, and proposed two more that have nicer properties, based on smoothing splines and distance correlation. In addition, two more indexes for measuring dependence have been proposed in the machine learning literature, based on information criteria, maximal and total information coefficient (MIC and TIC) [@Reshef1518], with computationally more efficient versions (MIC_e, TIC_e) [@JMLRv1715308]. These are related to original 1D projection pursuit indexes based on entropy (e.g. @huber85, @JS87). This provides seven current indexes for measuring dependence between two variables, and each is available in an R [@rref] package: scagnostics [@HWscagR], mbgraphic [@mbgraphic] and minerva [@minerva]. The projection pursuit guided tour is available in the R package, tourr [@tourr] provides optimisation routines, and visualization.

@@

The paper investigates the behaviour of these indexes. Section \ref{sec:construct} discusses index construction, and how they fit into the guided tour. Section \ref{sec:investigate} investigates the behaviour of the indexes, explored primarily using tour methods (@As85, @BCAH05, @CBCH94). The new guided tour with these indexes are applied to two examples from physics models (Section \ref{sec:phys}).

@@

- Standardizing each variable, to have mean 0 and variance 1, so that individual variable scales do not affect the result. Different variable scales are examinable without resorting to projection pursuit, so can be handled prior to searching through high dimensions.
- Sphering the high-dimensional data is often done to remove linear dependence. This is typically done using principal component analysis, and using the principal components as the variables passed to PP. If linear dependence is the only structure PP is not needed, and thus this is removed before PP so that the PPIs are not distracted by simple structure.
- Transform single variables to reduce skewness. It is marginal structure, visible in a single variable, which doesn't need a multivariate technique to reveal. Skewed distributions will inadvertently affect the PPIs, distracting the search for dependence. 
- Remove outliers, which may be an iterative process, to discover, identify and delete. Extreme values will likely affect PPI performance. Outliers can be examined on a case by case basis later.
- Possibly remove noise dimensions, which is also likely to be an iterative process. Directions where the distribution is purely noise make optimization of a PPI more difficult. If a variable is suspected to have little structure and relationship with other variables, conducting PP on the subset of variables without them may improve the efficiency of the search.
- Centering and scaling of the projected data, can be helpful visually. If the data has a small amount of non-normal distribution in some directions, the projected data can appear to wander around the plot window during a tour. It doesn't matter what the centre of the projected data is, so centering removes a wandering scatterplot. Less commonly, it may be useful to scale the projected data to standard values, which would be done to remove any linear dependence remaining in the data.

@@

\begin{table}[htp]
\caption{Summary of notation}
\centering
\begin{tabular}{|l|p{8.2cm}|} \hline
$p$ & be the number of variables in the data. For physics models the observable space is higher dimension, say $q$, and the data examined is the fitted model space, typically less than 10. \\
$d (=2) $ & projection dimension. For studying physics models, typically $d=2$ and this is the focus for the index definition. \\
$n$ & Number of observations, which for the physics models is the number of fitted models being examined and compared. \\
&\\
$\bm{X}_{n\times p} = (\bm{X}_1, ..., \bm{X}_p)$ & $n\times p$-dimensional data matrix, where variables $\bm{X}_j$ may be scaled or standardised, and $\bm{X}$ may be sphered\\
$\bm{Y}_{n\times 2} = (\bm{Y}_1, ..., \bm{Y}_2)$ & projected data matrix. where $Y_j = (\alpha_1 X_1, ..., \alpha_p X_p$ \\
$\bm{F}_{p\times 2} = (\bm{\alpha}_1, ..., \bm{\alpha}_2)$ & orthonomal projection matrix\\ 
&\\
$H$ & Convex hull \\
$A$ & Alpha hull \\\hline
\end{tabular}
\label{notation}
\end{table}

@@

* **scagnostics**: The first step to computing the scagnostics is that the bivariate data is binned, and scaled between [0,1] for calculations. The convex and alpha hulls, and the minimal spanning tree (MST), are computed. 
    + *convex*: The ratio of the area of alpha to convex hull, $I_{\bm{F}, convex}= \frac{area(A(Y))}{area(H(Y))}$. In this is the only measure where interesting projections will take low values, with a maximum of 1 if both areas are the same. Thus $1-c_{convex}$ is used.
    + *skinny*: The ratio of the perimeter to the area of the alpha hull, $I_{\bm{F}, skinny} = 1 -  \frac{\sqrt{4\pi area(A)}}{perimeter(A)}$, where the normalization is chosen such that $I_{\bm{F}, skinny} = 0$ for a full circle. Values close to 1 indicate a skinny polygon.
    + *stringy*: Based on the MST, $I_{\bm{F}, stringy} = \frac{diameter(MST)}{length(MST)}$ where the diameter is the longest connected path, and the length is the total length (sum of all edges). If the MST contains no branches $I_{\bm{F}, skinny} = 1$.

@@

## Optimization
Given a PPI we are confronted with the task of finding the maximum over all possible $d$ dimensional projections. One challenge is to avoid getting trapped in local maxima that are only a result of sampling fluctuations or a consequence of A noisy index function. \cite{posse95b} discusses the optimization, in particular that for most index functions and optimal results are too local, largely dependent on starting point. @f87 suggested a two-step procedure: the first step is using a large step size to find the approximate global maximum while stepping over pseudomaxima. A second step is then starting from the projection corresponding to the approximate maximum and employing a gradient directed optimization for the identification of the maximum. For exploring high-dimensional data, it can be interesting to observe local maxima as well as a global maximum, and thus a hybrid algorithm that still allows lingering but not being trapped by local maxima is ideal. In addition, being able to visually monitor the optimisation and see the optimal projection in the context of neighbouring projections is useful. This is provided by combining projection pursuit with the grand tour [@CBCH94]. The properties of a suitable optimization algorithm include monotonicity of the index value, a variable step-size to avoid overshooting and to increase the chance of reaching the nearest maximum, and a stopping criterion allowing to move out of a local maximum and into a new search region [@tourr]. A possible algorithm is inspired by simulated annealing and has been described in @lckl2005, this has been implemented in the \texttt{search\_better} and \texttt{search\_better\_random} search functions in the tourr package. The tourr package also provides the \texttt{search\_geodesic} function, which first selects a promising direction by comparing index values between a selected number of small random steps, and then optimises the function over the line along the geodesic in that direction considering projections up to $\pi/4$ away from the current view.

@@

A useful projection pursuit index needs to have several properties. This has been discussed in several seminal papers, e.g. @diaconis84, @huber85, @JS87, @posse95b, @hall89. The PPI should be  minimized by the normal distribution, because this is not interesting from a data exploration perspective. If all projections are normally distributed, good modelling tools already exist. A PPI should be approximately affine invariant, regardless how the projection is rotated the index value should be the same, and the scale of each variable shouldn't affect the index value. Interestingly the original index proposed by \cite{FT74} was not rotationally invariant. A consistent index means that small perturbations to the sample do not dramatically change the index value. This is particularly important to making optimization feasible, small angles between projections correspond to small perturbations of the sample, and thus should be small changes to index value. \cite{posse95b} suggests that indexes should be resistant to features in a tail of the distribution, but this is debatable because one departure from normality that is interesting to detect are anomalous observations. Some PPI are designed precisely for these reasons. Lastly, because we need to compute the PPI over many projections, it needs to be fast to compute. These form the basis of the criteria upon which the scagnostic indexes, and the several alternative indexes are examined, as explained below.

@@

Three families of data simulations are used for examining the behaviour of the index functions. Each generates  structure in two variables, with the remaining variables containing various types of noise. This is a very simple construction, because there is no need for projection pursuit to find the structure, one could simply use the PPIs on pairs of variables. However, it serves the purpose to also evaluate the PPIs. The three data families are explained below. In each set, $n$ is used for the number of points, $p$ is the number of dimensions, and $d=2$ is the projection dimension.

@@

Rescaling: first have cut-off at the respective value for the normal distribution, i.e. any value below 0.2 (0.8) is set to zero for cmass (holes) index, rescale remaining range to be between zero and one.

@@

XXX from caption, remove?:  For too rough binning (bins=10) we observe random fluctuations from the binning, with structure detection being obscured. With bins=20 we see somewhat smoother behaviour for the stringy index, especially when moving into the maximum, while the skinny index shows a dip in the final view, and larger bins=50 is preferred. For the convex index we also find bins=50 to be the better choice.

@@

## Binning sensitivity of MIC index
To examine the sensitivity of binning in the MIC PPI, the classic RANDU data [@Marsaglia68], available in R, is used. Binning is defined by $\delta$, where $B(n) = n^{\delta}$. Figure \ref{fig:randuEx} shows the best projection, index value and computing time obtained when optimizing the MIC index with values $\delta = 0.6, 0.7, 0.8$. With small $\delta$, less bins, the structure isn't visible, and with larger $\delta$ the structure is confounded with noise. It does appear that this parameter affects the performance of the MIC index.
